Sources of Measurement Bias in Modern CPUs

### 2014-01-01 paper.tex
Updated perf counters and content

\documentclass[a4paper,11pt,twocolumn,twoside]{article}
\usepackage{fullpage}
\usepackage{graphicx}
\usepackage{pgfplots}
\usepackage{pgfplotstable}
\usepackage{filecontents}
\usepackage{booktabs}
\usepackage{tabularx}
\usepackage{tikz}
\usepackage{listings}
\usepackage{authblk}
\usepackage{caption} % For margin top bottom

@@

\section{4K Address Aliasing}
An event known as “4K aliasing” can occur when the addresses of a store instruction followed by a load instruction differ by a multiple of 4 KiB.
On recent Intel architectures, only the last 12 address bits are used to determine if operations refer to the same address \cite{Intel:2012:OptimizationManual}.
This is a heuristic employed by the CPU to enable out of order execution of memory access operations.
Independent pairs of loads and stores where the last address bits differ can then be reordered and issued out of order.
Doing only a partial compare, \emph{false} dependencies can be detected when the addresses are equal $\bmod 4096$, as in the following example: 
\begin{lstlisting}[language={[x86masm]Assembler}]
mov %rax ($0x601010)
mov ($0x604010) %rbx
\end{lstlisting}
Situations like these causes the pipeline to stall and wait for the operations to finish in proper order.
The number of times this happens can be counted by the following performance counter:
\begin{description}
  \item{LD\_BLOCKS\_PARTIAL.ADDRESS\_ALIAS}\marginpar{\small{Evnt Mask \\ 0x07 0x01}} \hfill \\
  Counts the number of loads that have partial address match with preceding stores, causing the load to be reissued.
\end{description}
Performance implications of 4K aliasing is discussed to some extent in the optimization manual from Intel. 
There are even some concrete suggestions for how avoid aliasing cases to improve performance (Rule 8 and 9)~\cite{Intel:2012:OptimizationManual}.
In the following sections, we look at how address aliasing can explain some instances of measurement bias.

@@

\pgfplotstableread{resources/stack-offset.dat}{\stackoffsettable}
\begin{figure*}[t]
  \caption{Bias from environment size for program in Listing~\ref{lst:loopkernel}. Measure of average 100 cycle counts for 512 different environments, spread over two 4K periods. Compiled with no optimization.}
  \label{fig:envbias}
  \begin{tikzpicture}
    \begin{axis}[
        width = \textwidth,
        height = 6cm,
        font = \small,
        xlabel=Bytes added to environment,
        domain = 0:8192,
        xtick = {0,1024,...,8192},
        xmin = 0,
        xmax = 8192,
        legend pos = outer north east,
        cycle list name = exotic
      ]
      %\draw [<->, line width=0.8, color=red] (axis cs:0,8e5) -- (axis cs:4096,8e5) ;
      \addplot[ycomb] table[x expr = \thisrowno{0}*16, y = cycles:u] \stackoffsettable ;
      %\addplot table[x expr = \thisrowno{0}*16, y = r0107:u ] \stackoffsettable ;
      \addlegendentry{cycles:u} ;
      \addlegendentry{r0107:u} ;
    \end{axis}
  \end{tikzpicture}
\end{figure*}

@@

    % See page 453 in pgf manual
    \node [
      rectangle split, rectangle split parts=9, 
      rectangle split part fill={white, lightgray, white, lightgray, white, lightgray, white},
      draw, anchor=center, text width=2cm
    ] (m)  
      {
        \nodepart{one}
          stack
        \nodepart{three}
          mmap area
        \nodepart{five}
          heap
        \nodepart{seven}
          bss
        \nodepart{eight}
          data
        \nodepart{nine}
          text
      };

    \draw [decorate, decoration={brace, amplitude=5pt}] (m.south west) -- (m.six split west) 
      node [black, midway, xshift=-1.5cm, text width=2.0cm] 
        { Program code and static data } ;

    \node[below right] at (m.north east)
      { \scriptsize{0x7fff'ffffffff} } ;

    \node[above right] at (m.south east)
      { \scriptsize{0x400000} } ;

@@

There are clearly two worst cases, indicated by significant spikes around the 3100 and 7100 byte offsets.
We analyzed these cases by sampling an extensive set of performance counters and correlating them with cycle counts.
The results show a high number of resource stalls when the spikes occur, indicated by high values for performance counters RESOURCE\_STALLS.ANY, CYCLE\_ACTIVITY.CYCLES\_LDM\_PENDING, CYCLE\_ACTIVITY.CYCLE\_NO\_EXECUTE and RESOURCE\_STALLS.RS.
We also find that LD\_BLOCKS\_PARTIAL.ADDRESS\_ALIAS correlates almost perfectly with cycle count, which can explain the high number of resource stalls with false dependency from aliasing.

@@

\subsection{Heap Address Aliasing}
Address aliasing can be caused by conflicting pairs of load/store operations to any part of memory.
The previous section looked specifically at an example of collision between stack variables and static data.
Another scenario to consider is collisions in heap allocated memory.
In particular, code that operate on pairs of contiguous arrays can be vulnerable to 4K aliasing.
Acquiring dynamic memory at run time is usually done by calling malloc, which takes a number of bytes to allocate and returns a pointer to that area.
A typical malloc implementation uses two different strategies for how to allocate memory.
The \emph{heap} area in virtual memory is used for smaller allocations.
Programs are initially given a small heap area, which the allocator can increase by calling brk or sbrk.
For larger allocations, it is typically faster to use anonymous memory mapping by calling mmap, instead of growing the ``normal'' heap.
The heap starts at a low address close to the data section and grows towards higher addresses. 
Mmap allocates much closer to the stack.
While heap addresses can look like 0x16e30a0 or 0x1723020, pointers returned by mmap can for example be 0x7f0318a8f010 or 0x7f03105d2010.
This distinction is of course unimportant for application developers, as everything is conceptually the same ``heap''.
However, mmap has an interesting property in that allocations will always be page aligned.
The page size is 4096 bytes, meaning two pointers returned by mmap will \emph{always} alias.\footnote{libc's version of malloc adds 16 bytes of medatada at the beginning, thus every mmap'ed address ends with 0x010}

@@

\pgfplotstableread{bin/convolution.dat}{\convolutiontable}
\begin{figure}[t]
  \caption{Cycle and alias counts for different offsets between input and output arrays in Listing~\ref{lst:conv}. Offset 0 means equal 12 bit address suffix, which default behavior for mmap allocations. Array size $n=2^{20}$, compiled with -O3. }
  \label{fig:heapalias}
  \begin{tikzpicture}
    \begin{axis}[
        font = \small,
        xlabel=Offset in \lstinline!sizeof(float)! bytes,
        cycle list name=black white
      ]
      \addplot table[x expr = \thisrowno{0}, y = cycles:u] \convolutiontable ;
      \addplot table[x expr = \thisrowno{0}, y = r0107:u ] \convolutiontable ;
      \addlegendentry{cycles:u} ;
      \addlegendentry{r0107:u} ;
    \end{axis}
  \end{tikzpicture}
\end{figure}

@@

Figure \ref{fig:heapalias} shows ...

Using malloc to allocate input and output float arrays of size n, they will alias if mmap is used internally.
For sufficiently large values of n, this will always be the case.
Compiling with optimization O3 and n=$2^{20}$, a very large number of alias events and resource stalls are generated compared to cycle count.

\section{The Loop Stream Detector}
Typical software spends most of the time executing the same instructions repeatedly in loops, where the same instructions are fetched, decoded and executed many times. 
The Loop Stream Detector (LSD) is a front-end hardware optimization that is able to detect small software loops.
A cache of decoded instructions is kept and analyzed for cyclic chains in the instruction stream.
When a loop is detected, can be fetched directly from a queue of already decoded micro operations.
This enables some instruction decode logic in the processor front end to be temporarily disabled.
The LSD is a power saving optimization, but can also improve performance in cases where the decode pipeline is a bottleneck.

Utilization of the Loop Stream Detector can be measured by the following performance counter.
\begin{description}
  \item{LSD.UOPS}\marginpar{\small{Evnt Mask \\ 0xA8 0x01}} \hfill \\
  Counts the number of micro-operations delivered by the LSD.
\end{description}

@@

Some effort has been made to model properties of the LSD on previous Intel architectures, as an aid in low-level optimization. 
Hundt et. al.~\cite{Hundt:2011:MAO} looks at assembly rewrite rules for optimizing agains the Core 2 LSD.
For loops to be recognized, they have to fit within a set of restrictions imposed by the hardware implementation that tries to detect them. 
For instance, because the LSD cache is of fixed size there will be a maximum number of instructions a loop can contain for it to be recognized.
The capabilities of the LSD has improved with newer micro-architectures.
Some important limitations for Haswell's LSD are listed below.
\begin{itemize}
  \item Up to 12 chunk fetches of 32 B instruction memory.\footnote{The optimization manual actually states 8 chunk fetches for HTT enabled, and 11 for HTT disabled. Our experiments with a minimal test case shows that the LSD is active for loops spanning up to 12 chunks -- regardless of hyper threading.}
  \item Up to 28 micro-operations for HTT enabled, 56 for HTT disabled.
  \item Can not contain CALL or RET instructions
  \item Stack operations must match, i.e push must be matched with a pop.
\end{itemize}

@@

@@

\section{Conclusions}

>>>>

\documentclass[a4paper,10pt,twocolumn,twoside]{article}
\usepackage{fullpage}
\usepackage{graphicx}
\usepackage{pgfplots}
\usepackage{pgfplotstable}
\usepackage{filecontents}
\usepackage{booktabs}
\usepackage{tabularx}
\usepackage{tikz}
\usepackage{listings}
\usepackage{authblk}
\usepackage{caption} % For margin top bottom

@@

\section{4K Address Aliasing}
An event known as “4K aliasing” can occur when the addresses of a store instruction followed by a load instruction differ by a multiple of 4 KiB.
On recent Intel architectures, only the last 12 address bits are used to determine if operations refer to the same address \cite{Intel:2012:OptimizationManual}.
This is a heuristic employed by the CPU to enable out of order execution of memory access operations.
Independent pairs of loads and stores where the last address bits differ can then be reordered and issued out of order.
Doing only a partial compare, \emph{false} dependencies can be detected when the addresses are equal $\bmod 4096$, as in the following example: 
\begin{lstlisting}[language={[x86masm]Assembler}]
mov %rax ($0x601010)
mov ($0x604010) %rbx
\end{lstlisting}
Situations like these causes the pipeline to stall and wait for the operations to finish in proper order.
The number of times this happens can be counted by the following performance counter:
\begin{description}
  \item{LD\_BLOCKS\_PARTIAL.ADDRESS\_ALIAS}\marginpar{\vspace{0.0pt}\footnotesize\begin{tabular}{@{} l @{\hskip 2pt} l }Event & Mask \\ 0x07 & 0x01\end{tabular}} \hfill \\
  Counts the number of loads that have partial address match with preceding stores, causing the load to be reissued.
\end{description}
Performance implications of 4K aliasing is discussed to some extent in the optimization manual from Intel. 
There are even some concrete suggestions for how avoid aliasing cases to improve performance (Rule 8 and 9)~\cite{Intel:2012:OptimizationManual}.
In the following sections, we look at how address aliasing can explain some instances of measurement bias.

@@

\pgfplotstableread{bin/stack-offset.dat}{\stackoffsettable}
\begin{figure*}[t]
  \caption{Bias from environment size for program in Listing~\ref{lst:loopkernel}. Measure of average 100 cycle counts for 512 different environments, spread over two 4K periods. Compiled with no optimization.}
  \label{fig:envbias}
  \begin{tikzpicture}
    \begin{axis}[
        width = \textwidth,
        height = 6cm,
        font = \small,
        xlabel=Bytes added to environment,
        domain = 0:8192,
        xtick = {0,1024,...,8192},
        xmin = 0,
        xmax = 8192,
        legend pos = outer north east,
        cycle list name = exotic
      ]
      %\draw [<->, line width=0.8, color=red] (axis cs:0,8e5) -- (axis cs:4096,8e5) ;
      \addplot[ycomb] table[x expr = \thisrowno{0}*16, y = cycles:u] \stackoffsettable ;
      %\addplot table[x expr = \thisrowno{0}*16, y = r0107:u ] \stackoffsettable ;
      \addlegendentry{cycles:u} ;
      \addlegendentry{r0107:u} ;
    \end{axis}
  \end{tikzpicture}
\end{figure*}

@@

    % See page 453 in pgf manual
    \node [
      rectangle split, rectangle split parts=10, 
      rectangle split part fill={white, white, lightgray, white, lightgray, white, lightgray, white},
      draw, anchor=center, text width=2cm
    ] (m)
      {
        \nodepart{one}
          environment
        \nodepart{two}
          stack
        \nodepart{four}
          mmap area
        \nodepart{six}
          heap
        \nodepart{eight}
          bss
        \nodepart{nine}
          data
        \nodepart{ten}
          text
      };

    \draw [decorate, decoration={brace, amplitude=5pt}] (m.south west) -- (m.seven split west) 
      node [black, midway, xshift=-1.5cm, text width=2.0cm] 
        { Program code and static data } ;

    \node[right] at (m.north east)
      { \scriptsize{0x7fff'ffffffff} } ;

    \node[right] at (m.south east)
      { \scriptsize{0x400000} } ;

@@

There are clearly two worst cases, indicated by significant spikes around the 3100 and 7100 byte offsets.
We analyzed these cases by sampling an extensive set of performance counters and correlating them with cycle counts.
The results show a high number of resource stalls when the spikes occur, indicated by high values for performance counters \small{RESOURCE\_STALLS.ANY}, \small{CYCLE\_ACTIVITY.CYCLES\_LDM\_PENDING}, \small{CYCLE\_ACTIVITY.CYCLE\_NO\_EXECUTE} and \small{RESOURCE\_STALLS.RS}.
We also find that \small{LD\_BLOCKS\_PARTIAL.ADDRESS\_ALIAS} correlates almost perfectly with cycle count, which can explain the high number of resource stalls with false dependency from aliasing.

@@

\subsection{Heap Address Aliasing}
Address aliasing can be caused by conflicting pairs of load/store operations to any part of memory.
The previous section looked specifically at an example of collision between stack variables and static data.
Another scenario to consider is collisions in heap allocated memory.
In particular, code that operate on pairs of contiguous arrays can be vulnerable to 4K aliasing.
Acquiring dynamic memory at run time is usually done by calling malloc, which takes a number of bytes to allocate and returns a pointer to that area.
A typical malloc implementation uses two different strategies for how to allocate memory.
The \emph{heap} area in virtual memory is used for smaller allocations.
Programs are initially given a small heap area, which the allocator can increase by calling brk or sbrk.
For larger allocations, it is typically faster to use anonymous memory mapping by calling mmap, instead of growing the ``normal'' heap.
The heap starts at a low address close to the data section and grows towards higher addresses. 
Mmap allocates much closer to the stack.
While heap addresses can look like 0x16e30a0 or 0x1723020, pointers returned by mmap can for example be 0x7f0318a8f010 or 0x7f03105d2010.
This distinction is of course unimportant for application developers, as everything is conceptually the same ``heap''.
However, mmap has an interesting property in that allocations will always be page aligned.
The page size is 4096 bytes, meaning two pointers returned by mmap will \emph{always} alias.\footnote{libc's version of malloc adds 16 bytes of medatada at the beginning, therefore every mmap'ed address ends with 0x010}

@@

\pgfplotstableread{bin/convolution.dat}{\convolutiontable}
\begin{figure}[t]
  \caption{Cycle and alias counts for different offsets between input and output arrays in Listing~\ref{lst:conv}. Offset 0 means equal 12 bit address suffix, which is default behavior for mmap allocations. Array size $n=2^{20}$, compiled with -O3. }
  \label{fig:heapalias}
  \begin{tikzpicture}
    \begin{axis}[
        font = \small,
        xlabel=Relative offset in \lstinline!sizeof(float)! bytes,
        cycle list name=black white
      ]
      \addplot table[x expr = \thisrowno{0}, y = cycles:u] \convolutiontable ;
      \addplot table[x expr = \thisrowno{0}, y = r0107:u ] \convolutiontable ;
      \addlegendentry{cycles:u} ;
      \addlegendentry{r0107:u} ;
    \end{axis}
  \end{tikzpicture}
\end{figure}

@@

Figure \ref{fig:heapalias} shows how this function behaves when given input with different alignment.
The default behavior for large heap allocations is that both arrays alias.
To force different alignments, we allocate some extra bytes at the end and offset the output pointer argument.

\begin{lstlisting}
output = malloc(n * (sizeof(float) + offset));
conv(n, input, output + offset);
\end{lstlisting}

Listing \ref{lst:conv}, together with a small main method handling the allocation and offset, is compiled with optimization O3 and input size n=$2^{20}$.
Figure \ref{fig:heapalias} plots performance counter statistics for cycle count and 4K alias events over increasing amount of offset.
This is an example of how false dependencies from address aliasing can cause a significant performance hit.
Because large heap allocations always alias, the processor thinks memory accesses to input[i] potentially conflicts with output[i], stalling the pipeline.
By manually adjusting the alignment of one input buffer, cycle count can be reduced by more than 50\% (verify).
This is a speedup on top of compiler optimization O3, which is not able to account for aliasing effects.

It is perhaps unfair to assume the compiler can fix this when only given a generic optimization flag, as aliasing effects is (presumably) an Intel-specific issue.
You might also wamt to utilize the C99 keyword restrict to explicitly tell the compiler that the buffers are disjoint in memory.

\begin{lstlisting}[language=C]
void conv( int n,
            const float * restrict in,
            const float * restrict out );
\end{lstlisting}

Even with the updated signature compiled with -O3 -march=native, there is still room for improvement by inserting manual offsets.

\section{The Loop Stream Detector}
Typical software spends most of the time executing the same instructions repeatedly in loops, where the same instructions are fetched, decoded and executed many times. 
The Loop Stream Detector (LSD) is a front-end hardware optimization that is able to detect small software loops.
A cache of decoded instructions is kept and analyzed for cyclic chains in the instruction stream.
When a loop is detected, can be fetched directly from a queue of already decoded micro operations.
This enables some instruction decode logic in the processor front end to be temporarily disabled.
The LSD is a power saving optimization, but can also improve performance in cases where the decode pipeline is a bottleneck.
Utilization of the Loop Stream Detector can be measured by the following performance counter.
\begin{description}
  \item{LSD.UOPS\marginpar{\vspace{0.0pt}\footnotesize\begin{tabular}{@{} l @{\hskip 2pt} l }Event & Mask \\ 0xA8 & 0x01\end{tabular}}} \hfill \\
  Counts the number of micro-operations delivered by the Loop Stream Detector.
\end{description}

@@

Some effort has been made to model properties of the LSD on previous Intel architectures, as an aid in low-level optimization. 
Hundt et al.~\cite{Hundt:2011:MAO} looks at assembly rewrite rules for optimizing agains the Core 2 LSD.
For loops to be recognized, they have to match a set of restrictions imposed by the hardware implementation that tries to detect them. 
For instance, because the LSD cache is of fixed size there will be a maximum number of instructions a loop can contain for it to be recognized.
The capabilities of the LSD has improved with newer micro-architectures.
Some important limitations for Haswell's LSD are listed below.
\begin{itemize}
  \item Up to 12 chunk fetches of 32 B instruction memory.\footnote{The optimization manual actually states 8 chunk fetches for HTT enabled, and 11 for HTT disabled. Our experiments with a minimal test case shows that the LSD is active for loops spanning up to 12 chunks -- regardless of hyper threading.}
  \item Up to 28 micro-operations for HTT enabled, 56 for HTT disabled.
  \item Can not contain CALL or RET instructions
  \item Stack operations must match, i.e push must be matched with a pop.
\end{itemize}

@@

\begin{lstlisting}[language=C, escapeinside=``, caption={Example of bias from link order. Instruction alignment of the loop changes with ordering. }]
`loop.c`
#define B22(n) b += i*i*a + 1
volatile int i = 42;

int loop(int a) {
    register int b = i == 42 ? 1 : 0;
    do {
        // 10 bytes mov(6), test(2), jne(2)
        if (!i) B22(1); 
        if (!i) B22(2);
        if (!i) B22(3);
        if (!i) B22(4);
        if (!i) B22(5);
        if (!i) B22(6);
        if (!i) B22(7);
        if (!i) B22(8);
        if (!i) B22(9);
        if (!i) B22(10);
        if (!i) B22(11);
        // 27 bytes for increment and test
    } while (i++ < 0x12345678);
    return b;
}
`foo.c`
int foo(int n) {
    int a = 1;
    return a;
}
`main.c`
#include <stdio.h>
int main() {
    int i = loop(10);
    int f = foo(10);
    printf("Loop: %d\n", i);
    printf("Foo: %d\n", f);
    return 0;
}
\end{lstlisting}

@@

\section{Conclusions}
Any change to virtual memory layout of data can potentially introduce bias effects from address aliasing. 
We find that compilers, even when provided with architecture specific optimization flags, still generates code that suffers from aliasing. 
Typical heap allocators return page-aligned memory by default, which is the worst case with respect to aliasing for many algorithms.
Despite recommendations from Intel, we are not aware of any allocators that specifically addresses this issue.

The Loop Stream Detector can provide significant speedup, but subtle changes to code layout sometimes prevents this optimization to be utilized. 
Programmers rarely care, or even know, about intricate details such as the addresses and alignment of code. 
We expect, and probably rightfully so, that sensible compilers and linkers will organize code somewhat optimally. 
In cases where this is not true, weird effects that might be categorized as “bias” occurs. 
For developers, it is useful to be aware of the fact that any change to instruction addresses can potentially have a huge impact on performance.
